\ourTitle

### 2022-09-28 paper/manuscript.tex
paper edits

\def\ourAbstract{%
Predictive states for stochastic processes are a nonparametric and interpretable
construct with relevance across a multitude of modeling paradigms. Recent
progress on the self-supervised reconstruction of predictive states from
time-series data focused on the use of reproducing kernel Hilbert spaces. Here,
we examine how Wasserstein distances may be used to detect predictive
equivalences in symbolic data. We compute Wasserstein distances between
distributions over sequences (``predictions''), using a finite-dimensional
embedding of sequences based on the Cantor for the underlying geometry. We show
that exploratory data analysis using the resulting geometry via hierarchical
clustering and dimension reduction provides insight into the temporal structure
of processes ranging from the relatively simple (\emph{e.g.}, finite-state
hidden Markov models) to the very complex (\emph{e.g.}, infinite-state indexed
grammars).
}

@@

{\bf
\section*{Lead:}
Discovering the hidden structure of complex systems has been a widely-recognized
goal of nonlinear science for decades, originally starting with extracting
``geometry'' from a time series. The present results follow directly in this
long line of inquiry, especially that focusing on the causal states or
effective theories underlying time series generated by complex nonlinear
systems. It provides workable statistical estimation methods and dynamical
interpretations of hidden mechanisms. It establishes a new mathematical
foundation for causal inference of complex systems, partly by making
connections to modern machine learning. In this, its use of Cantor embeddings and Wasserstein distances complements recent work on reproducing-kernel Hilbert space representations.
}

@@

The following focuses on the first, as it is a crucial building block to
achieving the other two. Recent attempts to reconstruct the geometry of
predictive states embedded them in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do \cite{Loom21a}. Here, we achieve the same results using a
procedure that can be easily visualized and interpreted. This makes it suitable
as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}, while moving
further toward interpretable machine learning of structured processes.

@@

We refer back frequently to the following example processes of increasing
computational complexity:
\begin{enumerate}
      \setlength{\topsep}{-2pt}
      \setlength{\itemsep}{-2pt}
      \setlength{\parsep}{-2pt}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A
	typical example might look like $01100111101100011$.
\item The $\mathtt{a}^n \mathtt{b}^n$ \emph{process} can be generated by
	choosing a random integer $n\geq 1$ (we suppose via a Poisson process) and
	writing $n$ $\mathtt{a}$s followed by an equal number of $\mathtt{b}$s,
	and then repeating this procedure indefinitely. This results in sequences
	where any contiguous block of $\mathtt{a}$s is followed by a block of
	$\mathtt{b}$s of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot
	be generated by any finite hidden Markov chain, though it is a simple
	example of a probabilistic context-free language \cite{Hopc06a}. A typical
	example might look like $\mathtt{abaaabbbabaabb}$.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
\item The $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ \emph{process} is a
	probabilistic indexed language \cite{Hopc06a} that is analogous to
	$\mathtt{a}^n \mathtt{b}^n$ except after writing the blocks of
	$\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block of $\mathtt{c}$'s
	of length $n$. A typical example might look like
	$\mathtt{abcaaabbbcccabcaabbcc}$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}. A typical example might look like
	$11011101010111011101110101011101$---the fifth generation of the
	replacement rule starting from $0$.
\end{enumerate}

@@

The geometry of sequences is inherently self-similar. Given an infinite sequence
$\overrightarrow{x} = (x_1,x_2,\dots)$, we can split it into its leading word
$x_1 x_2 \dots x_L$ and a following sequence $\overrightarrow{x}_L =
(x_L,x_{L+1},\dots)$. That is, the space of sequences $\mathcal{X}^\mathbb{N}$
can be factored into $\mathcal{X}^L\times \mathcal{X}^\mathbb{N}$ for any $L$.
The fractal nature of sequence-space is encoded in the structure of its
\emph{product topology}.

We exploit this self-similarity in an interesting way by constructing a mapping
between sequence space and the celebrated \emph{Cantor set} (or one of its
generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
associate a unique integer between $0$ and $|\mathcal{X}|-1$ inclusive; call
this $J(x)$. Then, there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
[0,1]$ that maps every sequence to a positive real number:
\begin{align*}
  C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
  ~.
\end{align*}
For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
sequence to a point the traditional Cantor set fractal. For a finite sequence of
length $L$, truncate the sum at $k=L$.

Remarkably, the embedding $C$ has the property that for any continuous function
$f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
is continuous on $\mathcal{X}^\mathbb{N}$. Further, if $F$ is continuous on
$\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
Thus, the embedding $C$ respects the basic structure of the product topology
\cite{Kurk03a}.

@@

Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
structure that may or may not be desired. Associating each symbol $x$ with an
integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
embeddings that do not have this. However, they come at the cost of increased
computational complexity when determining the Wasserstein distance. So, for now,
we assume that ordinal artifacts are either desired or sufficiently tolerable
as to not outweigh the computational benefits of working in one dimension.
\begin{algorithm}[t]
  \SetKwData{UnqPasts}{UnqPasts}\SetKwData{Cantors}{Cantors}\SetKwData{Wass}{Wass}
  \SetKwFunction{Wasserstein}{Wasserstein}\SetKwFunction{Append}{append}
  \SetKwFunction{Index}{index} \SetKwFunction{Length}{length}
  \SetKwFunction{Matrix}{Matrix}
  \SetKwFunction{CantorWasserstein}{CantorWasserstein}
  \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

@@

For example, by examining the even process' Cantor fractal, one notices that
there are effectively only $2$ distinct predictive states---every vertical
column is just one of two types. This corresponds with the $2$ states of the
hidden Markov model that generates the even process.

@@

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by:
\begin{align*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
   ~,
\end{align*}
where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
It is the minimal cost to ``shift'' the probability mass from one distribution
to match the other's shape.

$W(\mu,\nu)$ is the solution to a constrained linear optimization. As a
function of distributions, $W(\mu,\nu)$ is continuous with respect to
convergence in distribution. In fact, convergence under the Wasserstein
distance is equivalent to convergence in distribution on compact spaces. This
makes $W(\mu,\nu)$ ideal for measuring geometry between predictive states,
since empirical estimates of these are known to converge in distribution.

When $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
respectively be the cumulative distributions functions of $\mu$ and $\nu$.
Then:
\begin{align*}
  W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
  ~.
\end{align*}
This closed-form solution is considerably faster to compute than the linear
optimization required for arbitrary metric spaces. Since the Cantor embedding
embeds the space of sequences directly into $[0,1]$, we can directly employ
this formula.

@@

Figure \ref{fig:mds} plots the first two dimensions of a multidimensional
scaling (MDS) decomposition \cite{Borg05a} for the even and
$\mathtt{a}^n\mathtt{b}^n$ processes. Clusters are colored in the same manner
as in Fig. \ref{fig:hclust} and labeled by the specific pattern that
distinguishes the pasts in some of the clusters. Note that the clusters and
labels are directly drawn from Fig. \ref{fig:hclust} for reference. They are
not the result of the MDS algorithm itself. However, interactive plotting
approaches may allow for similar exploration from these decompositions without
the need for prior clustering.

The even process, as in all other cases seen thus far, has two dominant
prediction clusters. These correspond to the predictive states that result
from seeing an even-sized block of $1$s (or, equivalently, no $1$s) and that
result from seeing an odd-sized block of $1$s. The lone cluster in the middle
corresponds to a transient state induced by seeing all $1$s and then a $0$.
The latter then synchronizes to right cluster.

@@

>>>>

\def\ourAbstract{%
Predictive states for stochastic processes are a nonparametric and interpretable
construct with relevance across a multitude of modeling paradigms. Recent
progress on the self-supervised reconstruction of predictive states from
time-series data focused on the use of reproducing kernel Hilbert spaces. Here,
we examine how Wasserstein distances may be used to detect predictive
equivalences in symbolic data. We compute Wasserstein distances between
distributions over sequences (``predictions''), using a finite-dimensional
embedding of sequences based on the Cantor set for the underlying geometry. We show
that exploratory data analysis using the resulting geometry via hierarchical
clustering and dimension reduction provides insight into the temporal structure
of processes ranging from the relatively simple (\emph{e.g.}, finite-state
hidden Markov models) to the very complex (\emph{e.g.}, infinite-state indexed
grammars).
}

@@

{\bf
\section*{Lead:}
Discovering the hidden structure of complex systems has been a widely-recognized
goal of nonlinear science for decades, originally starting with extracting
``geometry'' from a time series. The present results follow directly in this
long line of inquiry, especially that focusing on the causal states or
effective theories underlying time series generated by complex nonlinear
systems. It provides workable statistical estimation methods and dynamical
interpretations of hidden mechanisms. It expands upon a new mathematical
foundation for causal inference of complex systems, partly by making
connections to modern machine learning. In this, its use of Cantor embeddings and Wasserstein distances complements recent work on reproducing-kernel Hilbert space representations.
}

@@

The following focuses on the first, as it is a crucial building block to
achieving the other two. Recent attempts to reconstruct the geometry of
predictive states embedded them in reproducing kernel Hilbert spaces (RKHS)
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do \cite{Loom21a}. Here, we achieve the same results using a
procedure that can be easily visualized and interpreted. This makes it suitable
as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}, while moving
further toward interpretable machine learning of structured processes.

@@

We refer back frequently to the following example processes of increasing
computational complexity:
\begin{enumerate}
      \setlength{\topsep}{-2pt}
      \setlength{\itemsep}{-2pt}
      \setlength{\parsep}{-2pt}
\item The \emph{fair coin process} is simply generated by flipping a coin repeatedly and
writing down a $0$ for every tail and $1$ for every head. There is no memory in
this process, and all pasts will map to the same predictive state. We will
discuss it occasionally as a counterpoint to the more complex examples below.
We also provide code and figures for this example in our GitHub repository
\cite{repo}. 
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A
	typical example might look like $01100111101100011$.
\item The $\mathtt{a}^n \mathtt{b}^n$ \emph{process} can be generated by
	choosing a random integer $n\geq 1$ (we suppose via a Poisson process) and
	writing $n$ $\mathtt{a}$s followed by an equal number of $\mathtt{b}$s,
	and then repeating this procedure indefinitely. This results in sequences
	where any contiguous block of $\mathtt{a}$s is followed by a block of
	$\mathtt{b}$s of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot
	be generated by any finite hidden Markov chain, though it is a simple
	example of a probabilistic context-free language \cite{Hopc06a}. A typical
	example might look like $\mathtt{abaaabbbabaabb}$.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a finite sequence of $A$s (such as $AAA$). Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
	The sequence generation process does not ``end'' until all non-terminals
	(that is, $A$'s, $B$'s and $C$'s) have been replaced by terminal symbols.
	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$ which
	results from the substitutions
	\begin{align*}
		AAA &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}
		\ C\ \mathtt{;}\ C\ \mathtt{;}\\
		&\mapsto C\mathtt{+x;}
		\mathtt{f(}B\mathtt{);}\mathtt{f(}B\mathtt{);}\\
		&\mapsto \mathtt{x+x;f(}B\ \mathtt{+}\ C\mathtt{);f(}C\mathtt{);}\\
		&\mapsto \mathtt{x+x;f(}C\ \mathtt{+f(}C\mathtt{));f(f(}C\mathtt{));}\\
		&\mapsto \mathtt{x+x;f(x+f(x));f(f(x));}
	\end{align*}
\item The $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ \emph{process} is a
	probabilistic indexed language \cite{Hopc06a} that is analogous to
	$\mathtt{a}^n \mathtt{b}^n$ except after writing the blocks of
	$\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block of $\mathtt{c}$'s
	of length $n$. A typical example might look like
	$\mathtt{abcaaabbbcccabcaabbcc}$. Though it looks very similar to the
	$\mathtt{a}^n \mathtt{b}^n$ process, it is in a formally distinct complexity
	class, as $\mathtt{a}^n \mathtt{b}^n$ may be generated with a simple stack
	automaton while $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ requires nested stacks.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}. A typical example might look like
	$11011101010111011101110101011101$---the fifth generation of the
	replacement rule starting from $0$.
\end{enumerate}

Though our examples are abstract, they represent real structures which may arise
in practice.

Examples 3 through 5, in particular, involve long-range
correlations in their sequences which arise from structures which, upon
``opening,'' must be ``closed'' an arbitrary time later. Such structures are
present in human language as well as in genetic sequences. In the latter case,
this long-range correlation exists due to the folding of proteins which allows
distant codons to interact with one another. Detecting and modeling these
folding structures requires
more complex models than hidden Markov models, such as the context-free grammars
in examples 3 and 4 \cite{Saka94a}. 
Example 6 represents the general problem of recovering knowledge of a dynamical
system of which only coarse-grained data is available.

@@

The geometry of sequences is inherently self-similar. Given an infinite sequence
$\overrightarrow{x} = (x_1,x_2,\dots)$, we can split it into its leading word
$x_1 x_2 \dots x_L$ and a following sequence $\overrightarrow{x}_L =
(x_{L+1},x_{L+2},\dots)$. That is, the space of sequences $\mathcal{X}^\mathbb{N}$
can be factored into $\mathcal{X}^L\times \mathcal{X}^\mathbb{N}$ for any $L$;
much like a fractal, then, $\mathcal{X}^\mathbb{N}$ is comprised of copies of itself.
The fractal nature of sequence-space is encoded in the structure of its
\emph{product topology}, in which the closeness of two sequences is measured by
the number of sites at which their values match. In other words, the
neighborhoods of the product topology are generated by these smaller copies of
$\mathcal{X}^\mathbb{N}$.

We can visualize this self-similarity in an interesting way by constructing a mapping
between sequence space and the celebrated \emph{Cantor set} (or one of its
generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
associate a unique integer between $0$ and $|\mathcal{X}|-1$ inclusive; call
this $J(x)$. Then, there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
[0,1]$ that maps every sequence to a positive real number:
\begin{align*}
  C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
  ~.
\end{align*}
For instance, suppose that $|\mathcal{X}|=2$ has two elements; then the mapping
$C$ looks like
\begin{align*}
	C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{3^k}
	~.
  \end{align*}
The range of this mapping, taken over all binary sequences, matches the
traditional Cantor set fractal, which is attained by removing the middle third
from the unit interval, and repeating this process on all remaining intervals
\emph{ad infinitum}. When $\mathcal{X}$ has more than two elements, the
resulting range is a generalized form of the Cantor set, where instead of
removing the middle (second) third, we remove the second and fourth quintiles ($|\mathcal{X}|=3$), or
second, fourth and sixth heptiles ($|\mathcal{X}|=4$), and so on.
For a finite sequence of
length $L$, we truncate the sum at $k=L$, resulting in a finite-depth
approximation to the Cantor set.

Why use this somewhat convoluted construction for an embedding? The reason is
that the embedding $C$ preserves the basic structure of the product topology on
sequences \cite{Kurk03a}. Imagine, for instance, that we chose a more direct
map, such as mapping a binary sequence to the corresponding binary number
(\emph{e.g.} $010110\dots \mapsto 0.010110\dots$): then dissimilar sequences
such as $0111\dots$ and $1000\dots$ would in fact map to the same real number.
The padding added by removed intervals in the Cantor set means we can map distinct sequences to distinct real
numbers, while keeping similar sequences close together.

@@

Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
structure that may or may not be desired. Associating each symbol $x$ with an
integer $J(x)$ endows an ordinal structure on the set $\mathcal{X}$. That is,
the specific mapping from symbols to integers is arbitrary, and the way the
symbols are ordered in the Cantor embedding---for instance, which symbol gets
assigned to the middle segment instead of the end segments when
$|\mathcal{X}|=3$---can affect the results of the visualization and downstream
processing. Higher-dimension embeddings may be contrived which do not impose an
ordering on the symbols. However, they come at the cost of increased
computational complexity when determining the Wasserstein distance. So, for now,
we assume that ordinal artifacts are either desired or sufficiently tolerable as
to not outweigh the computational benefits of working in one dimension.
\begin{algorithm}[t]
  \SetKwData{UnqPasts}{UnqPasts}\SetKwData{Cantors}{Cantors}\SetKwData{Wass}{Wass}
  \SetKwFunction{Wasserstein}{Wasserstein}\SetKwFunction{Append}{append}
  \SetKwFunction{Index}{index} \SetKwFunction{Length}{length}
  \SetKwFunction{Matrix}{Matrix}
  \SetKwFunction{CantorWasserstein}{CantorWasserstein}
  \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

@@

For example, by examining the even process' Cantor fractal, one notices that
there are effectively only $2$ distinct predictive states---every vertical
column is just one of two types. This corresponds with the $2$ states of the
hidden Markov model that generates the even process. By comparison, if we had
plotted the fair coin's Cantor fractal, we would have found it to be perfectly
symmetric, with every vertical column identical.

@@

This provides some intuition for how to effectively choose a distance metric for
predictive states. Metrics such as the total variation distance or
divergence-based distance metrics are too coarse; a small shift in the support
of the distributions can result in dramatic changes in the distances between the
distributions. Instead, we want a distance metric for distributions that
quantifies the differences between the supports. The Wasserstein metric, also known as
the earth-mover's distance, provides this feature \cite{Pane19a}. As the name ``earth-mover's''
suggests, we may imagine the probability distributions as masses of dirt, and
quantify the difference between two distribution as the cost of shifting the
dirt around to transform one distribution to another. This way, if two
distributions have similar but non-overlapping supports, the Wasserstein metric
will be small.

Formally, we can define the Wasserstein metric as follows. Given two measures
$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by:
\begin{align*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
   ~,
\end{align*}
where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
It is the minimal cost to ``shift'' the probability mass from one distribution
to match the other's shape.

$W(\mu,\nu)$ is the solution to a constrained linear optimization (since the
objective function and constraints are linear functions of $\pi$). This can
become computationally costly, scaling as $O(n^3 \log n)$ in the number of
samples $n$ \cite{Pele09a}.
However, when $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
respectively be the cumulative distributions functions of $\mu$ and $\nu$.
Then:
\begin{align*}
  W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
  ~.
\end{align*}
This closed-form solution is considerably faster to compute than the linear
optimization required for arbitrary metric spaces. Since the Cantor embedding
embeds the space of sequences directly into $[0,1]$, we can directly employ
this formula.

Topologically, the Wasserstein distance replicates the topology of convergence
in distribution, which means that algorithms which construct predictive states using the
Wasserstein distance will converge.

@@

Figure \ref{fig:mds} plots the first two dimensions of a multidimensional
scaling (MDS) decomposition \cite{Borg05a} for the even and
$\mathtt{a}^n\mathtt{b}^n$ processes. MDS extracts coordinates so that the
Euclidean distance between points in MDS coordinates matches the Wasserstein
distance between the distributions being embedded. Clusters are colored in the
same manner as in Fig. \ref{fig:hclust} and labeled by the specific pattern that
distinguishes the pasts in some of the clusters. Note that the clusters and
labels are directly drawn from Fig. \ref{fig:hclust} for reference. They are not
the result of the MDS algorithm itself. However, interactive plotting approaches
may allow for similar exploration from these decompositions without the need for
prior clustering.

The even process, as in all other cases seen thus far, has two dominant
prediction clusters. These correspond to the predictive states that result
from seeing an even-sized block of $1$s (or, equivalently, no $1$s) and that
result from seeing an odd-sized block of $1$s. The lone cluster in the middle
corresponds to a transient state induced by seeing all $1$s and then a $0$.
The latter then synchronizes to right cluster. By comparison, the fair coin's
MDS embedding would have a single cluster since every past predicts the same
future.

@@

We also note that the direct reconstruction of the logistic map in this case is
possible due to the choice of partition to transform the continuous variable $y$
into the discrete variable $x$. The partition we chose is a \emph{generating
partition} of the dynamical system \cite{Kurk03a}, which is a special kind of
partition which preserves many of the measure-theoretic and topological features
of the original system. Non-generating partitions would not yield the original
logistic map if our methods are applied to them; however, nor would they if any
other method were applied, as a non-generating partition represents an
irreversible loss of information. In this case predictive state reconstruction
would still recover the minimal sufficient statistics of the past sequence for
future prediction (see, for instance, the discussion in \cite{Crut98d}).
